Estimating changes in temperature distributions in a large ensemble of climate simulations using quantile regression

### 2018-05-13 paper_revision.tex
address minor comments and specifics

\abstract{
Understanding future changes in extreme temperature events is critically important because of their potential societal cost, but inherently challenging in a transient climate. A single model simulation is generally insufficient to empirically characterize the statistical properties of underlying processes. Ensembles of repeated simulations (with different initial conditions providing independence) not only alleviate this data insufficiency but also allow new approaches for characterizing changes in extremes. We present here a new method of using ensembles that allows characterizing changes in temperature distributions using a continuous representation of seasonality rather than breaking the dataset into seasonal blocks. That is, we assume that temperature distributions evolve smoothly both day-to-day over an annual cycle and year-to-year over longer secular trends. To demonstrate our method's utility, we analyze an ensemble of 50 simulations of the Community Earth System Model (CESM) under a scenario of increasing radiative forcing to 2100, focusing on North America. The results both confirm aspects of climate system behavior known from previous studies and also elucidate new features.Confirming results include that daily temperature bulk variability generally decreases  in wintertime in the continental mid- and high-latitudes ($> 40^{\circ}$). New results include that in these same wintertime distributions, the low tails ``stick'', i.e.\ experience a lesser reduction in variability, producing a more negative skew. Although the examples above concern temperature only, the technique is sufficiently general that it can be used to generate precise estimates of distribution changes in a broad range of climate variables by exploiting the power of ensembles.}

@@

Climate timeseries have generally been assumed to be separable into two components: randomness inherent in the underlying physical processes, which we call natural variability, and forced secular trends that follow from increasing concentrations of greenhouse gases.
Recently, the degree to which that natural variability may itself be changing has received significant scientific interest \citep[e.g.][]{trenberth2011attribution, donat2012shifting,deser2012communication, thompson2015quantifying, kay2015community}. Potential changes in climate extremes, because of their heightened societal impacts, are of special concern \citep[e.g.][]{davison1990models, stott2004human, chavez2005generalized, eastoe2009modelling, otto2012reconciling, swain2014extraordinary, singh2014severe, trenberth2015attribution, diffenbaugh2015anthropogenic, huang2015changes, jalbert2017spatiotemporal}. 
However, detecting and understanding such changes on timescales relevant to human activities is inherently challenging and generally data-limited.
%identifying changes in climate variability related to human activities is inherently challenging. 
%The responses in the climate system are generally assumed to be separable into two components: 
The long equilibration time of the climate system means that on the timescales of interest to human society, the climate state will be evolving (or `transient'), so that its statistical properties are not stationary. Fully characterizing this evolving natural  variability of rare events is intrinsically challenging due to the limited amount of data in either observations or simulations generally available. Studies of future climate extremes often employ statistical extreme value theory to make inferences about rare events with modest amounts of data.
% XX some flipping of sentence order here, and made the key word "develop methods" instead of just "study" - EJM.

n this work, we study the entire distribution of temperatures in a transient climate, including rare events, by employing quantile regression on an ensemble of simulations of an identical forcing scenario from a single climate model.
% Each simulation has a different set of initial conditions to reflect the natural variability of the system. 
%Given sufficient care with choice of initial conditions, this approach can ensure that each simulation is statistically independent in terms of its natural variability.) 
Given sufficient care in choosing different initial conditions, such a large ensemble will reflect the natural variability of the system, since each simulation will be be statistically independent in terms of its natural variability. 
The increased data provided by multiple simulations %XX allows instead of admits?
then admits more confident statements about changes in the statistical behavior of the system %XX compared to 
than can be made with a single simulation. While the use of ensembles is a relatively recent development, it is growing rapidly \citep[e.g.][]{deser2012uncertainty, deser2012communication,deser2014projecting,fischer2014detection, kay2015community,sriver2015effects, rodgers2015emergence, hagos2016projection}. \citet{deser2012uncertainty}, \citet{deser2012communication} and \citet{fischer2014detection} in particular discuss how ensembles help in distinguishing internal climate variability from anthropogenic effects and allow more accurate estimates for the forced model response. There remains however considerable room for additional development of statistical methods that exploit the benefits of ensembles.

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Ensembles of multiple simulations offer at least three advantages that are currently under-exploited. The most important such advantage is that the increased data volume allows examining the entire distribution of a climate variable.
 Studies of climate variability to date are generally divided between those that address the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics}, and those that address its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, generally via extreme value theory. %In part the division has arisen from the data scarcity imposed by particular datasets.  %---- END CC
%High-resolution dynamically downscaled model studies \citep{XX,XX,XX} could in principle involve multiple runs, but their computational demands generally limit them to at most a few decades of present and future simulations. 
% XX Those studies 
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. 
When studying model projections using ensembles, the large amount of data at each location allows us to accurately estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the purpose of this analysis, we compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Over the region of analysis, the model adequately reproduces overall geospatial patterns in temperature mean, standard deviation, and skewness, though with some quantitative discrepancies. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability somewhat, and produces excessively cold winter temperatures in the Arctic. The resulting exaggerated temperature gradients contribute to excess variability and excessively negative skew in the northern mid-latitudes.
(Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}.) %% ejm changed from cubes to cube.
 To build intuition around our data, we will show throughout this work results for the three locations \textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}, which have representative but different temperature distributions. See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

% CC BEGIN
 As an example of a typical model fit, we show in Figure \ref{Estimate} the seasonal cycle in CESM daily temperatures for three locations, along with estimates of low, median and high quantiles. We show here data from 1850 to demonstrate the seasonal fit rather than that of the long-term trend. All locations show strong seasonal differences in variance that are well-represented by our smooth estimates. Relevant features that are captured include an asymmetrical seasonal cycle in all locations; a clear left skewness in wintertime in all three locations (although most pronounced in the higher-latitude \textbf{a} and \textbf{b}); and a distinct springtime shoulder in the higher-latitude locations. %All three sites show that the transition from cold to warm temperatures in spring looks different from the warm to cold transition in fall, suggesting 
These characteristics show the benefit of explicitly modeling seasonal variations, since analyzing seasonal averages, and perhaps even monthly averages, would throw out meaningful details.

@@

To facilitate comparison to previous studies, we first perform a preliminary analysis where we replicate more standard methods. That is, we examine changes in the aggregate distribution of temperatures over multi-week and multi-month intervals, before we show results from our new approach that calculates responses for individual days. Even the standard analysis readily shows that temperature distributions in the CESM ensemble change markedly over the RCP 8.5 scenario (Figures \ref{location_distributions}, \ref{meanStdSkew_Winter}, and \ref{meanStdSkew_Summer}, which compare the “initial” and “final” time windows 1850-1864 and 2086-2100). Means uniformly shift warmer, but the shapes of the distributions also change in complex ways. Figure \ref{location_distributions} shows initial and final distributions in our example locations for aggregated 15-day periods in winter and summer. In at least two of the three depicted locations, it is clear that the distributions are becoming narrower, although quantifying exactly how the tails are changing requires more sophisticated techniques.
%difficult to quantify exactly how the tails are changing.

The changes in distributions also exhibit complex spatial patterns. Figures \ref{meanStdSkew_Winter} and \ref{meanStdSkew_Summer} show patterns of change in mean, standard deviation, and skewness using 3-month seasonal blocks (DJF and JJA). We see the expected strong decrease in variance in winter over land, especially in the northern mid-latitudes. By contrast, summer variance changes are much smaller and differ in sign in different locations. Temperature skewness, i.e. the asymmetry of the distribution, shows strong changes in winter over land in a dipole pattern. Winter temperature distributions are in all time periods negatively skewed throughout most of the domain, but in the north (including locations \textbf{a} and \textbf{b}), they become more negatively skewed in the future, while in the south (including location \textbf{c}), they become more symmetric.  Summer skewness changes are again smaller and with less spatial coherence, other than the strong transitions in the Southern Great Plains and in Mexico/Central America, where 
skewness in temperature distributions actually changes sign.

@@

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the 0.25th and 0.75th quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently than does the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodoogy allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{p, i}$ and $\Delta q_{p, f}$ at the $p^{th}$ percentile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{p, f} - \Delta q_{p, i}}{\Delta q_{p, i}}.
\label{eq:relnorm}
\end{equation}
 Because our method reveals the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.

Results  show that tail changes can indeed differ strongly from changes in the bulk of the distribution.  In wintertime (Figure \ref{winter_summer_var_avg}, top row), in much of the northrn mid-latitudes (including locations \textbf{a} and \textbf{b}), low tails change in a way that contributes to more a more negative skew. Low tail variability contracts less than does the IQR, while high tail variability contracts more strongly. (High tails would contribute to more negative winter skew predomimantly in the Hudson's Bay region, where the model shows distinct bias.) In summertime (Figure \ref{winter_summer_var_avg}, bottom row), the high tail dominates the transition to positive skew in the Southern Great Plains region (including location \textbf{c}).
%to the production of negative skew in southern Mexico/Central America. Bulk variability increases, but variability increases more in the low tail and less in the high tail than in the IQR.   
See Supplementary Online Material for additional related figures that provide context.  Figure \ref{winter_summer_var_avg_big} shows the same analysis over a larger region, and Figures \ref{winter_summer_var_mn} and \ref{winter_summer_var_mx} repeat Figure \ref{winter_summer_var_avg} but for daily maximum and minimum temperatures, respectively, rather than for daily means. 
%Summer variability changes more in the low tail and the IQR than in the high tail. Substantial widening of the temperature distribution's IQR and low tail is seen in Central America, in turn causing a more negative skew as seen in Figure \ref{meanStdSkew_Summer}. Information about which tail change causes change in skewness is a detail made possible through our proposed method.

@@

To clarify the relative contributions of high and low tails to skewness changes, we 
also examine evolving temperature variability in the bulk and tails as a function of seasonality as well as long term change. 
Figure \ref{dq_vs_day} shows absolute variability changes 
for the three example locations \textbf{a}, \textbf{b}, and \textbf{c}
estimated using our quantile model, and for fractional changes see Supplementary Online Material Figure \ref{dq_vs_day_norm}.  %Again, each locations' temperature distributions are shown in Figure \ref{location_distributions} for summer and winter. 
  %  shows variability both in the bulk and in the tails for each day of the year 
Colored lines show different years in the 1850-2100 simulations, and enveloping dashed lines represent $90\%$ confidence intervals. See Appendix \ref{app:model}\ref{app:uq} for details about uncertainty quantification.
Horizontal variations are the seasonal evolution of variability; vertical offsets between lines are long-term changes.
In all locations, wintertime skewness changes are driven by the relative changes in IQR and low tails. In the higher-latitude locations \textbf{a} and \textbf{b}, more negative winter skew results because the IQR contracts even more strongly than does the low tail variability. In other words, the low tails ``stick''.  In the lower-latitude location \textbf{c}, more positive winter skew results because the IQR barely changes while the low tail variability contracts strongly.

@@

The complexity of the relationships in Figure \ref{dq_vs_day} also shows how misleading it may be to use a three-month block to represent a season. 
 With the detailed information that results from our approach, we can see just how complex even the present day variability profiles are. While all three locations show higher bulk variability (IQR) in winter than summer, the transition from winter to summer happens more quickly in some locations than at others, more abruptly in the northernmost location \textbf{a} and more gradually in \textbf{c}. Low-tail variability seasonal transitions are even sharper than those of IQR in \textbf{a} and \textbf{b}, but more gradual in \textbf{c}.
%also has very different profiles between each location. Location \textbf{a} hits peak low-tail variability during the shoulder seasons (spring and fall), while location \textbf{b} shows a more gradual low tail variability rise toward winter. Location \textbf{c} shows less temporal change in the IQR compared with the other locations. Overall, 
In contrast, high-tail variability is more seasonally constant overall than low-tail variability. 
Through these examples, we see how our method offers detailed information about changes in variability across seasons and annual change, usually unavailable when analyzing each season separately.

>>>>

\abstract{
Understanding future changes in extreme temperature events in a transient climate is inherently challenging. A single model simulation is generally insufficient to empirically characterize the statistical properties of the underlying physical processes governing the climate. Ensembles of repeated simulations (with different initial conditions providing independence) not only alleviate this lack of data but also allow additional new approaches for characterizing changes in extremes. We present here one such new approach, using ensembles that allow characterizing changes in temperature distributions using a continuous representation of seasonality rather than breaking the dataset into seasonal blocks. That is, we assume that temperature distributions evolve smoothly both day-to-day over an annual cycle and year-to-year over longer secular trends. To demonstrate our method's utility, we analyze an ensemble of 50 simulations of the Community Earth System Model (CESM) under a scenario of increasing radiative forcing to 2100, focusing on North America. The results both confirm aspects of climate system behavior known from previous studies and also elucidate new features. Confirming results include that daily temperature bulk variability generally decreases  in wintertime in the continental mid- and high-latitudes ($> 40^{\circ}$). One new result is that these same wintertime distributions, the low tails ``stick'', i.e.\ experience a lesser reduction in variability, producing a more negative skew. Although the examples above concern temperature only, the technique is sufficiently general that it can be used to generate precise estimates of distribution changes in a broad range of climate variables by exploiting the power of ensembles.}

@@

Climate timeseries have generally been assumed to be separable into two components: randomness inherent in the underlying physical processes, which we call natural variability, and forced secular trends that follow from increasing concentrations of greenhouse gases.
Recently, the degree to which natural variability may itself be changing has received significant scientific interest \citep[e.g.][]{trenberth2011attribution, donat2012shifting,deser2012communication, thompson2015quantifying, kay2015community}. Potential changes in climate extremes, because of their heightened societal impacts, are of special concern \citep[e.g.][]{davison1990models, stott2004human, chavez2005generalized, eastoe2009modelling, otto2012reconciling, swain2014extraordinary, singh2014severe, trenberth2015attribution, diffenbaugh2015anthropogenic, huang2015changes, jalbert2017spatiotemporal}. 
However, detecting and understanding such changes on timescales relevant to human activities is inherently challenging and generally data-limited.
%identifying changes in climate variability related to human activities is inherently challenging. 
%The responses in the climate system are generally assumed to be separable into two components: 
The long equilibration time of the climate system means that on the timescales of interest to human society, the climate state will be evolving (or `transient'), so that its statistical properties are not stationary. Fully characterizing this evolving natural  variability of rare events is intrinsically challenging due to the limited amount of available observations or simulation data. Studies of future climate extremes often employ statistical extreme value theory to make inferences about rare events with modest amounts of data.
% XX some flipping of sentence order here, and made the key word "develop methods" instead of just "study" - EJM.

In this work, we study the entire distribution of temperatures in a transient climate, including rare events, by employing quantile regression on an ensemble of simulations of an identical forcing scenario from a single climate model.
% Each simulation has a different set of initial conditions to reflect the natural variability of the system. 
%Given sufficient care with choice of initial conditions, this approach can ensure that each simulation is statistically independent in terms of its natural variability.) 
Given sufficient care in choosing different initial conditions, such a large ensemble will reflect the natural variability of the system, since each simulation will be statistically independent in terms of its natural variability. 
The increased data provided by multiple simulations %XX allows instead of admits?
then admits more confident statements about changes in the statistical behavior of the system %XX compared to 
than can be made with a single simulation. While the use of ensembles is a relatively recent development, it is growing rapidly \citep[e.g.][]{deser2012uncertainty, deser2012communication,deser2014projecting,fischer2014detection, kay2015community,sriver2015effects, rodgers2015emergence, hagos2016projection}. \citet{deser2012uncertainty}, \citet{deser2012communication} and \citet{fischer2014detection} in particular discuss how ensembles help in distinguishing internal climate variability from anthropogenic effects and allow more accurate estimates for the forced model response.

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Ensembles of multiple simulations offer at least three advantages that are currently under-exploited. The most obvious advantage is that the increased data volume allows examining the entire distribution of a climate variable.
 Studies of climate variability to date are generally divided between those that address the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics}, and those that address its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, generally via extreme value theory. %In part the division has arisen from the data scarcity imposed by particular datasets.  %---- END CC
%High-resolution dynamically downscaled model studies \citep{XX,XX,XX} could in principle involve multiple runs, but their computational demands generally limit them to at most a few decades of present and future simulations. 
% XX Those studies 
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. 
When studying model projections using ensembles, the large amount of data at each location allows us to accurately estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the purpose of this analysis, we compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability in some places, and produces excessively cold winter temperatures in the Arctic. The resulting exaggerated temperature gradients contribute to excess variability and excessively negative skew in the northern mid-latitudes.
(Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}.) %% ejm changed from cubes to cube.
Throughout this work, we will show in-depth analysis from three locations with distinct temperature distributions to highlight our proposed method (\textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}). See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

% CC BEGIN
 As an example of a typical model fit, we show in Figure \ref{Estimate} the seasonal cycle in CESM daily temperatures for three locations, along with estimates of low, median and high quantiles. We show here data from 1850 to demonstrate the seasonal fit rather than that of the long-term trend. All locations show strong seasonal differences in variance that are well-represented by our smooth estimates. Relevant features that are captured include an asymmetrical seasonal cycle in all locations; a clear left skewness in wintertime in all three locations (although most pronounced in the higher-latitude \textbf{a} and \textbf{b}); and a distinct springtime shoulder in the higher-latitude locations. These characteristics show the benefit of explicitly modeling seasonal variations as smoothly varying functions as opposed to a set of four constant functions changing value with season. Nuances like the decrease in winter temperature spread (variability) from early to late winter would not be captured by a piecewise constant model.

@@

To facilitate comparison to previous studies, we first perform a preliminary analysis where we replicate more standard methods. That is, we examine changes in the aggregate distribution of temperatures over multi-week and multi-month intervals, before we show results from our new approach that calculates responses for individual days. Even the standard analysis readily shows that temperature distributions in the CESM ensemble change over the RCP 8.5 scenario (Figures \ref{location_distributions}, \ref{meanStdSkew_Winter}, and \ref{meanStdSkew_Summer}, which compare the “initial” and “final” time windows 1850-1864 and 2086-2100). Means uniformly shift warmer, but the shapes of the distributions also change in terms of variance and skewness. Figure \ref{location_distributions} shows initial and final distributions in our example locations for aggregated 15-day periods in winter and summer. In at least two of the three depicted locations, it is clear that the distributions are becoming narrower, although quantifying exactly how the tails are changing requires a quantification of the tail size and shape.
%difficult to quantify exactly how the tails are changing.

Regarding the spatial characteristics of temperature distributions, we see the expected strong decrease in variance in winter over land, especially in the northern mid-latitudes (Figures \ref{meanStdSkew_Winter} and \ref{meanStdSkew_Summer}). By contrast, summer variance changes are much smaller and differ in sign in different locations. Temperature skewness, i.e. the asymmetry of the distribution, shows strong changes in winter over land in a dipole pattern. Winter temperature distributions are in all time periods negatively skewed throughout most of the domain, but in the north (including locations \textbf{a} and \textbf{b}), they become more negatively skewed in the future, while in the south (including location \textbf{c}), they become more symmetric.  Summer skewness changes are again smaller and with less spatial coherence, other than the strong transitions in the Southern Great Plains and in Mexico/Central America, where 
skewness in temperature distributions actually changes sign.

@@

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the 0.25th and 0.75th quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently than does the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{p, i}$ and $\Delta q_{p, f}$ at the $p^{th}$ percentile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{p, f} - \Delta q_{p, i}}{\Delta q_{p, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.

Results  show that tail changes can indeed differ strongly from changes in the bulk of the distribution.  In wintertime (Figure \ref{winter_summer_var_avg}, top row), in much of the northrn mid-latitudes (including locations \textbf{a} and \textbf{b}), low tails change in a way that contributes to more a more negative skew. Low tail variability contracts less than does the IQR, while high tail variability contracts more strongly. (High tails would contribute to more negative winter skew predomimantly in the Hudson Bay region, where the model shows distinct bias.) In summertime (Figure \ref{winter_summer_var_avg}, bottom row), the high tail dominates the transition to positive skew in the Southern Great Plains region (including location \textbf{c}).
%to the production of negative skew in southern Mexico/Central America. Bulk variability increases, but variability increases more in the low tail and less in the high tail than in the IQR.   
% See Supplementary Online Material for additional related figures that provide context.  Figure \ref{winter_summer_var_avg_big} shows the same analysis over a larger region, and Figures \ref{winter_summer_var_mn} and \ref{winter_summer_var_mx} repeat Figure \ref{winter_summer_var_avg} but for daily maximum and minimum temperatures, respectively, rather than for daily means. 
%Summer variability changes more in the low tail and the IQR than in the high tail. Substantial widening of the temperature distribution's IQR and low tail is seen in Central America, in turn causing a more negative skew as seen in Figure \ref{meanStdSkew_Summer}. Information about which tail change causes change in skewness is a detail made possible through our proposed method.

@@

To clarify the relative contributions of high and low tails to skewness changes, we 
also examine evolving temperature variability in the bulk and tails as a function of seasonality as well as long term change. 
Figure \ref{dq_vs_day} shows absolute variability changes 
for the three example locations \textbf{a}, \textbf{b}, and \textbf{c}
estimated using our quantile model, and for fractional changes see Supplementary Online Material Figure \ref{dq_vs_day_norm}. The uncertainty around our estimates is quantified by resampling the original simulations (with replacement) and recomputing the estimates using this new set of simulations (see Appendix \ref{app:model}\ref{app:uq} for details). In all locations, wintertime skewness changes are driven by the relative changes in IQR and low tails. In the higher-latitude locations \textbf{a} and \textbf{b}, more negative winter skew results because the IQR contracts even more strongly than does the low tail variability. In other words, the low tails ``stick''.  In the lower-latitude location \textbf{c}, more positive winter skew results because the IQR barely changes while the low tail variability contracts strongly.

@@

The complexity of the relationships in Figure \ref{dq_vs_day} also shows how misleading it may be to use a three-month block to represent a season. While all three locations show IQR in winter than summer, the transition from winter to summer happens more quickly in some locations than at others, more abruptly in the northernmost location \textbf{a} and more gradually in \textbf{c}. Low-tail variability seasonal transitions are even sharper than those of IQR in \textbf{a} and \textbf{b}, but more gradual in \textbf{c}.
%also has very different profiles between each location. Location \textbf{a} hits peak low-tail variability during the shoulder seasons (spring and fall), while location \textbf{b} shows a more gradual low tail variability rise toward winter. Location \textbf{c} shows less temporal change in the IQR compared with the other locations. Overall, 
In contrast, high-tail variability is more seasonally constant overall than low-tail variability. 
Through these examples, we see how our method offers detailed information about changes in variability across seasons and annual change, usually unavailable when analyzing each season separately.
